---
title: "p8105_hw2_qz2527.Rmd"
author: "kindle zhang"
date: "2023-10-01"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1

First, clean the data in `pols-month.csv`. Use `separate()` to break up the variable
`mon` into integer variables `year`, `month`, and `day`; replace month number with month
name; create a `president` variable taking values `gop` and `dem`, and remove `prez_dem`
and `prez_gop`; and remove the day variable.

```{r}

month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

my_data_1 = read.csv("data_file/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into = c("year", "month", "day"), "-") |> 
  mutate(
    year = as.numeric(year),
    month = as.numeric(month),
    day = as.numeric(day)) |> 
  mutate(
    month = case_match(
        month,
        01 ~ "January",
        02 ~ "February",
        03 ~ "March",
        04 ~ "April",
        05 ~ "May",
        06 ~ "June",
        07 ~ "July",
        08 ~ "August",
        09 ~ "September",
        10 ~ "October",
        11 ~ "November",
        12 ~ "December",
      )
  ) |> 
  mutate(president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |> 
  select(year, month, everything(), -day, -starts_with("prez"))
```

Second, clean the data in `snp.csv` using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that `year` and `month` are the leading columns.

```{r clean data}
my_data_2 = 
  read_csv(
    "data_file/snp.csv",
    col_types = cols(date = col_date(format = "%m/%d/%y"))) |> 
  janitor::clean_names() |> 
  separate(date, into = c("year","month_num", "day"), convert = TRUE) |> 
  mutate(year = if_else(year>2023, year-100, year)) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, close)
  
```
Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r}
my_data_3 = 
  read_csv("./data_file/unemployment.csv") |> 
  rename(year = Year)|> 
  pivot_longer(
    Jan:Dec,
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```
Join the datasets by merging `snp` into `pols`, and merging `unemployment` into the result.

```{r}
data_new = 
  left_join(my_data_1, my_data_2) |> 
  left_join(x = _, y = my_data_3)
  
str(data_new)
```
Let's talk about the 538 data_new. The `my_data_1` data has `r nrow(my_data_1)` observations and `r ncol(my_data_1)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r my_data_1 |> pull(year) |> min()` to `r my_data_2 |> pull(year) |> max()`. It also tells us whether the sitting president was a democrat or republican. The `my_data_2` data has `r nrow(my_data_2)` observations and `r ncol(my_data_2)` variables, ranging from years `r my_data_2 |> pull(year) |> min()` to `r my_data_2 |> pull(year) |> max()`. The `my_data_3` data has `r nrow(my_data_3)` observations and `r ncol(my_data_3)` variables ranging from years `r my_data_3 |> pull(year) |> min()` to `r my_data_3 |> pull(year) |> max()`. In Januarys in or after 1975 in which a democrat was president, the **average unemployment rate was `r filter(data_new, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`**.  The average unemployment rate over the same time period in which a republican was president was `r filter(data_new, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.

## Problem 2